---
title: "**Homework 4 Classification**"
author: "Enes Ata Oruç"
date: "29.01.2021"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  
  The aim of this homework is to compare the performance of penalized regression approaches, decision trees, support vector machines and tree-based ensembles.
We are asked to find 4 datasets for a classification task from different domains. This report is created for BankChurners Dataset.

### Credit Card Churn Data:  (https://www.kaggle.com/sakshigoyal7/credit-card-customers)
  This dataset contains 740 observations and 21 features extracted from Cardiotocogram exams. Target column determined by the owners of the data. The target column is Attrition_Flag.
2- Existing Customer
1- Churned Customer

  

```{r, message=FALSE, warning=FALSE}
library(caTools)
library(randomForest)
library(ROCR)
library(tidyverse)
library(data.table)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(mlbench)
library(MLmetrics)
library(dplyr)
library(caret)
library(readr)
```

## 1. Penalized Regression Approach: ##

Altough I couldn't apply PRA properly (As with other datasets.) I wanted to share the code I used. I am open for feedbacks for this approach...

```{r pra}
# Read data
churn <- fread("BankChurners.csv")

##
##
# Not Working Properly

library(glmnet)
churn$Attrition_Flag <- as.factor(churn$Attrition_Flag)
smp_size_pra <- floor((2/3) * nrow(churn))

# set the seed to make your partition reproducible
set.seed(1234)
train_ind_pra <- sample(seq_len(nrow(churn)), size = smp_size_pra)

train_pra <- churn[train_ind_pra, ]
test_pra <- churn[-train_ind_pra, ]


#churn_mat_2 <- data.matrix(train_pra)
#cvfit=cv.glmnet(churn_mat_2, train_pra$Attrition_Flag,family='gaussian',nfolds=10)
#cvfit
#lambda_min <- cvfit$lambda.min
#lambda_min
# Lasso with Train Set
#lassofit <- glmnet(x=churn_mat_2, y = train_pra$Attrition_Flag, alpha = 1,lambda = lambda_min)
#lasso_train_predicted <-as.numeric(predict(lassofit, newx=data.matrix(train_pra)))

# RMSE for Train Lasso
#caret::RMSE(lasso_train_predicted, train_pra$Attrition_Flag)

# Lasso with Test Set
#lasso_test_predicted <-as.numeric(predict(lassofit, newx=data.matrix(test_pra)))
# RMSE for Test Lasso
#caret::RMSE(lasso_test_predicted, test_pra$Attrition_Flag)

```

## 2. Decision Tree Approach: ##

```{r churn, message=FALSE, warning=FALSE}
# 75% of the sample size churn
churn$Attrition_Flag <- as.numeric(churn$Attrition_Flag)
smp_size_dt <- floor((2/3) * nrow(churn))

# set the seed to make your partition reproducible
set.seed(123)
train_ind_dt <- sample(seq_len(nrow(churn)), size = smp_size_dt)

train_dt_churn <- churn[train_ind_dt, ]
test_dt_churn <- churn[-train_ind_dt, ]

#method is "class" because I want to perform classification
mytree_churn <- rpart(
  Attrition_Flag ~ ., 
  data = train_dt_churn, 
  method = "class"
)

# plot mytree
fancyRpartPlot(mytree_churn, caption = NULL)

printcp(mytree_churn) # display the results 
plotcp(mytree_churn) # visualize cross-validation results 
summary(mytree_churn) # detailed summary of splits

#prediction
predictions <- predict(mytree_churn, test_dt_churn, type = "class")
predictions = as.numeric(predictions)
# Prediction measures
table(test_dt_churn$Attrition_Flag, predictions)
#RMSE Value for performance test
RMSE(test_dt_churn$Attrition_Flag,predictions)
```

## 3. Random Forest Approach on churneeism Dataset: ##

```{r}
#Random Forest Approach
churn$Attrition_Flag = as.numeric(churn$Attrition_Flag)
#Test/train split
smp_size <- floor((2/3) * nrow(churn))

# set the seed to make your partition reproducible
set.seed(1234)

train_ind <- sample(seq_len(nrow(churn)), size = smp_size)
train_rf <- churn[train_ind, ]
test_rf <- churn[-train_ind, ]
# randomize data
random_index <- sample(1:nrow(train_rf), nrow(train_rf))
random_rf_train <- train_rf[random_index, ]

# Fit random forest: model
# Set seed
set.seed(33)
# Fit a model
model <- train(Attrition_Flag~.,
               data = random_rf_train,
               method = "ranger",
               trControl = trainControl(method = "cv", number = 5)
                             )
# Let's check the model
model
fit <- predict(model, newdata = test_rf)
fit_model <- train(Attrition_Flag~.,
               data = test_rf,
               method = "ranger",
               trControl = trainControl(method = "cv", number = 5)
                            )
fit_model
test_rf$Attrition_Flag <- as.numeric(test_rf$Attrition_Flag)
# RMSE for measure error
RMSE(test_rf$Attrition_Flag,fit)
```

## 4. Stochastic Gradient Approach: ##

```{r sga, message=FALSE, warning=FALSE}
library(ggplot2)
library(corrplot)
library(caret)
library(xgboost)

churn = churn[,-c(1,22,23)]
churn %>% select(where(is.numeric)) %>% as.matrix() %>%
cor() %>% corrplot(method = "number", type="lower")

dmy <- dummyVars(" ~ .", data = churn, fullRank = T)
BankChurners_transformed <- data.frame(predict(dmy, newdata = churn))

BankChurners_transformed.y = BankChurners_transformed[,1]
BankChurners_transformed.x = BankChurners_transformed[,-1]

trainIndex = sample(1:length(BankChurners_transformed.y), length(BankChurners_transformed.y)*.8)

#creating training data set by selecting the output row values
train.x = BankChurners_transformed.x[trainIndex,]
train.y = BankChurners_transformed.y[trainIndex]

#creating test data set by not selecting the output row values
test.x = BankChurners_transformed.x[-trainIndex,]
test.y = BankChurners_transformed.y[-trainIndex]

train.data.x <- as.data.frame(lapply(train.x, as.numeric))
test.data.x <- as.data.frame(lapply(test.x, as.numeric))

# convert data to xgboost format
train.xgb <- xgb.DMatrix(data = data.matrix(train.data.x), label = train.y)
test.xgb <- xgb.DMatrix(data = data.matrix(test.data.x), label = test.y)
params <- list (
                eta = 0.4,
                max_depth = 10,
                min_child_weight = 0.5,
                subsample = 1,
                colsample_bytree = 1,
                objective = "multi:softmax",
                eval_metric = "merror",
                num_class = 2
)
xgb.model <- xgb.train(params, train.xgb, nrounds = 10)

# predict values for test data
xgb.predict <- predict(xgb.model, test.xgb)
xgb.cm = confusionMatrix(as.factor(xgb.predict), as.factor(test.y))
print(xgb.cm)


```

## Compare Results ##
Penalized Regression Approach
```{r compare pra, message=FALSE, warning=FALSE}
# RMSE for Test Lasso
#caret::RMSE(lasso_test_predicted, test_pra$Attrition_Flag)
```
Decision Tree Approach
```{r compare DT, message=FALSE, warning=FALSE}
#RMSE Value for performance test DT
RMSE(test_dt_churn$Attrition_Flag,predictions)
```
Random Forest
```{r compare rf, message=FALSE, warning=FALSE}
# RMSE Value for performance test
RMSE(test_rf$Attrition_Flag,fit)
```
Stochastic Gradient Boosting
```{r compare sgb, message=FALSE, warning=FALSE}
# results
print(xgb.cm)
```

  When we looked at the RMSE values I believe there is some miscalculation happened in lasso approach because its value doesn't seem realistic. This lasso situation also happened on different datasets. But I tuned the other 3 approaches. Because of this meaningful assumptions, I believe they gave more reliable results. 

  So, when we consider the other approaches Gradient Boosting gave the best classification result on churn data.Its accuracy is also really high. There was 2 different classes in Target colum (Attrition_Flag). This target column determined by the researches that provided this dataset. 

2- Existing Customer
1- Churned Customer










