---
title: "**Homework 4 Classification**"
author: "Enes Ata Oruç"
date: "29.01.2021"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  The aim of this homework is to compare the performance of penalized regression approaches, decision trees, support vector machines and tree-based ensembles.
We are asked to find 4 datasets for a classification task from different domains. This report is created for Student Performance Dataset.

### Student Performance Data: ###
  This dataset contains 649 observations and 33 features about student performance in Porteguese Language. This dataset also mixed with numeric and categorical features and was modeled under binary/five-level classification and regression tasks. 
Student Performance Data in Porteguese Class: (https://archive.ics.uci.edu/ml/datasets/student%2Bperformance)
  
  There is also a paper published about this dataset. There will be Pass and Fail condition
according to student grades.

Related paper: http://www3.dsi.uminho.pt/pcortez/student.pdf

```{r,message=FALSE, error=FALSE}
# important libraries
library(tidyverse)
library(data.table)
library(rpart)
library(rattle)
library(rpart.plot)
library(RColorBrewer)
library(mlbench)
library(MLmetrics)
library(dplyr)
library(caret)
library(readr)

```
  
  Creating the Letter grade column based on overall grade(G3). Pass/Fail decision made by the boundaries 
stated in the published paper for this dataset. If G3>10 Pass(1), if <10 Fail(0).
Numbers and Letter representations:
1 -> Pass
2 -> Fail


  G1,G2 and G3 removed from the data because they directly correlated with the Letter grades. So, overfiting prevented.
```{r,message=FALSE, error=FALSE}
stud_dt <- fread("student.csv")
stud_dt$Record = ifelse(stud_dt$G3 >=10, 1,2)
stud_dt$Record = as.numeric(stud_dt$Record)
stud_dt$G1 <- NULL
stud_dt$G2 <- NULL
stud_dt$G3 <- NULL
stud_dt$Letter <- NULL

```

## 1. Penalized Regression Approach: ##

```{r}

library(glmnet)
stud_dt$Record <- as.numeric(stud_dt$Record)
smp_size_pra <- floor((2/3) * nrow(stud_dt))

# set the seed to make your partition reproducible
set.seed(1234)
train_ind_pra <- sample(seq_len(nrow(stud_dt)), size = smp_size_pra)

train_pra <- stud_dt[train_ind_pra, ]
test_pra <- stud_dt[-train_ind_pra, ]


health_mat_2 <- data.matrix(train_pra)
cvfit=cv.glmnet(health_mat_2, train_pra$Record,family='gaussian',nfolds=10)
cvfit
lambda_min <- cvfit$lambda.min
lambda_min
# Lasso with Train Set
lassofit <- glmnet(x=health_mat_2, y = train_pra$Record, alpha = 1,lambda = lambda_min)
lasso_train_predicted <-as.numeric(predict(lassofit, newx=data.matrix(train_pra)))

# RMSE for Train Lasso
caret::RMSE(lasso_train_predicted, train_pra$Record)

# Lasso with Test Set
lasso_test_predicted <-as.numeric(predict(lassofit, newx=data.matrix(test_pra)))
# RMSE for Test Lasso
caret::RMSE(lasso_test_predicted, test_pra$Record)

```

## 2. Decision Tree Approach: ##

```{r stud_dt, message=FALSE, warning=FALSE}
# 2/3 of the sample size stud_dt
stud_dt$Record <- as.numeric(stud_dt$Record)
smp_size_dt <- floor((2/3) * nrow(stud_dt))

# set the seed to make your partition reproducible
set.seed(123)
train_ind_dt <- sample(seq_len(nrow(stud_dt)), size = smp_size_dt)

train_dt_stud <- stud_dt[train_ind_dt, ]
test_dt_stud <- stud_dt[-train_ind_dt, ]

#method is "class" because I want to perform classification
mytree_stud <- rpart(
  Record ~ ., 
  data = train_dt_stud, 
  method = "class"
)

# plot mytree
fancyRpartPlot(mytree_stud, caption = NULL)

printcp(mytree_stud) # display the results 
plotcp(mytree_stud) # visualize cross-validation results 
summary(mytree_stud) # detailed summary of splits

#prediction
predictions <- predict(mytree_stud, test_dt_stud, type = "class")
predictions = as.numeric(predictions)
# Prediction measures
table(test_dt_stud$Record, predictions)
#RMSE Value for performance test
RMSE(test_dt_stud$Record,predictions)
```

## 3. Random Forest Approach ##

```{r}
#Random Forest Approach
stud_dt$Record = as.factor(stud_dt$Record)
#Test/train split
smp_size <- floor((2/3) * nrow(stud_dt))

# set the seed to make your partition reproducible
set.seed(1234)
train_ind <- sample(seq_len(nrow(stud_dt)), size = smp_size)
train_rf <- stud_dt[train_ind, ]
test_rf <- stud_dt[-train_ind, ]
x <- train_rf[,1:30]
y <- train_rf[,31] 
#RF Implement
control <- trainControl(method="repeatedcv", number=10, repeats=3)
seed <- 5
metric <- "Accuracy"
set.seed(1234)
mtry <- sqrt(ncol(x))
tunegrid <- expand.grid(.mtry=mtry)
rf_default <- train(Record~., data=train_rf, method="rf", metric=metric, tuneGrid=tunegrid, trControl=control)
print(rf_default)
# accuracy calculation of prediction
predictForest_2 = predict(rf_default, newdata = test_rf)
table(test_rf$Record, predictForest_2)

test_rf$Record<-as.numeric(test_rf$Record)
#RMSE Value for performance test
predictForest_2 <- as.numeric(predictForest_2)
RMSE(test_rf$Record,predictForest_2)
```

## 4. Stochastic Gradient Approach: ##

```{r sga, message=FALSE, warning=FALSE}
library(gbm)

stud_dt$Record = as.numeric(stud_dt$Record)
stud_dt$age = as.factor(stud_dt$age)
stud_dt$sex = as.factor(stud_dt$sex)
stud_dt$school = as.factor(stud_dt$school)
stud_dt$address = as.factor(stud_dt$address)
stud_dt$famsize = as.factor(stud_dt$famsize)
stud_dt$Pstatus = as.factor(stud_dt$Pstatus)
stud_dt$Mjob = as.factor(stud_dt$Mjob)
stud_dt$Fjob = as.factor(stud_dt$Fjob)
stud_dt$reason = as.factor(stud_dt$reason)
stud_dt$guardian = as.factor(stud_dt$guardian)
stud_dt$schoolsup = as.factor(stud_dt$schoolsup)
stud_dt$famsup = as.factor(stud_dt$famsup)
stud_dt$paid = as.factor(stud_dt$paid)
stud_dt$activities = as.factor(stud_dt$activities)
stud_dt$nursery = as.factor(stud_dt$nursery)
stud_dt$higher = as.factor(stud_dt$higher)
stud_dt$internet = as.factor(stud_dt$internet)
stud_dt$romantic = as.factor(stud_dt$romantic)
#Test/train split
smp_size <- floor((2/3) * nrow(stud_dt))

# set the seed to make your partition reproducible
set.seed(1234)

train_ind <- sample(seq_len(nrow(stud_dt)), size = smp_size)
train_sgb <- stud_dt[train_ind, ]
test_sgb <- stud_dt[-train_ind, ]
# randomize data
random_index <- sample(1:nrow(train_sgb), nrow(train_sgb))
random_sgb_train <- train_sgb[random_index, ]

# modify hyperparameter grid
hyper_grid <- expand.grid(
  shrinkage = c(.01, .05, .1),
  interaction.depth = c(3, 5, 7),
  n.minobsinnode = c(5, 7, 10),
  bag.fraction = c(.65, .8, 1), 
  optimal_trees = 0,               # a place to dump results
  min_RMSE = 0                     # a place to dump results
)

# total number of combinations
nrow(hyper_grid)
## [1] 81

# grid search 
for(i in 1:nrow(hyper_grid)) {
  
  # reproducibility
  set.seed(123)
  
  # train model
  gbm.tune <- gbm(formula = Record ~ .,
    distribution = "gaussian",
    data = random_sgb_train,
    n.trees = 600,
    interaction.depth = hyper_grid$interaction.depth[i],
    shrinkage = hyper_grid$shrinkage[i],
    n.minobsinnode = hyper_grid$n.minobsinnode[i],
    bag.fraction = hyper_grid$bag.fraction[i],
    train.fraction = .75,
    n.cores = NULL, # will use all cores by default
    verbose = FALSE
  )
  
  # add min training error and trees to grid
  hyper_grid$optimal_trees[i] <- which.min(gbm.tune$valid.error)
  hyper_grid$min_RMSE[i] <- sqrt(min(gbm.tune$valid.error))
}

hyper_grid %>% 
  dplyr::arrange(min_RMSE) %>%
  head(10)

# for reproducibility
set.seed(123)

# train GBM model
gbm.fit.final <- gbm(formula = Record ~ .,
  distribution = "gaussian",
  data = train_sgb,
  n.trees = 483,
  interaction.depth = 5,
  shrinkage = 0.1,
  n.minobsinnode = 5,
  bag.fraction = .65, 
  train.fraction = 1,
  n.cores = NULL, # will use all cores by default
  verbose = FALSE
  )  
# predict values for test data
pred <- predict(gbm.fit.final, n.trees = gbm.fit.final$n.trees, test_sgb)

# results
caret::RMSE(pred, test_sgb$Record)
```

## Compare Results ##
Penalized Regression Approach
```{r compare pra, message=FALSE, warning=FALSE}
# RMSE for Test Lasso
caret::RMSE(lasso_test_predicted, test_pra$Record)
```
Decision Tree Approach
```{r compare DT, message=FALSE, warning=FALSE}
#RMSE Value for performance test DT
RMSE(test_dt_stud$Record,predictions)
```
Random Forest
```{r compare rf, message=FALSE, warning=FALSE}
# RMSE Value for performance test
predictForest_2 <- as.numeric(predictForest_2)
RMSE(test_rf$Record,predictForest_2)
```
Stochastic Gradient Boosting
```{r compare sgb, message=FALSE, warning=FALSE}
# results
caret::RMSE(pred, test_sgb$Record)
```


  When we looked at the RMSE values I believe there is some miscalculation happened in lasso approach because its value doesn't seem realistic. This lasso situation also happened on different datasets. But I tuned the other 3 approaches. Because of this meaningful assumptions, I believe they gave more reliable results. 

  So, when we consider the other approaches Random Forest gave the best classification result on Student performance data with lowest RMSE value. But Stochastic Gradient Boosting also performed similar to Random Forest. Targer determined as Pass or Fail class. This target column determined by the researches that provided this dataset. 
